# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

Unzip the data, load it and attach the data set.

```{r}
unzip("activity.zip")
library(foreign)
stepsdata <- read.csv("activity.csv")
attach(stepsdata)
```

## What is mean total number of steps taken per day?

To calculate the mean steps per day we use the aggregate function.

```{r}
daystep <- aggregate(steps, by=list(date),FUN=sum)[,2]
mean(daystep, na.rm=TRUE)
median(daystep, na.rm=TRUE)
```

A fuller view of the distribution is provided by the histogram.

```{r}
hist(daystep,breaks=10)
```

## What is the average daily activity pattern?

The within-day pattern is also interesting. To look at this we aggregate by the 5-minute observation unit.

```{r}
slotstep  <- aggregate(steps ~ interval, FUN=sum)[,2]
```

To plot this as a time series, we need a single day's worth of 5-minute
observation units, which we also convert into decimal time (i.e., 1230
becomes 12.5) to avoid jumps on the x-axis.

```{r}
intday <- interval[1:288]
int <- floor(intday/100) + (intday %% 100)/60
plot(cbind(int,slotstep), type="l")
```

The busiest time of day is in the morning. The single busiest slot (and its value) is given by this code:

```{r}
intday[slotstep==max(slotstep)]
max(slotstep)
```

## Imputing missing values

A substantial amount of cases are missing data on the number of steps (but not on the other variable):

```{r}
sum(is.na(date))
sum(is.na(interval))
sum(is.na(steps))
```

We replace the missing values with the average observed value for the
corresponding time-slot. First we create a vector of mean values per
time-slot, and then replicate it to match the 61 days of observation:

```{r}
meanstep  <- aggregate(steps ~ interval, FUN=mean)[,2]
longmean <- rep(meanstep, 61)
```

We now create a new vector, stepsi, as a copy of steps, replacing
missing values with the corresponding mean:
```{r}
stepsi <-  steps
stepsi[is.na(stepsi)] <- longmean[is.na(stepsi)]
```

A histogram of the imputed data:
```{r}
daystepi <- aggregate(stepsi, by=list(date),FUN=sum)[,2]
hist(daystepi,breaks=10)
```

If we compare means and medians of the raw and imputed data we see that
imputing by mean values makes no difference to the daily means, but
changes the median slightly:
```{r}
mean(daystep, na.rm=TRUE)
mean(daystepi)
median(daystep, na.rm=TRUE)
median(daystepi)
```

## Are there differences in activity patterns between weekdays and weekends?

What are the differences between weekdays and weekends? To address this
we calculate the average per 5-min time-unit across weekdays and
weekends separately

First we identify weekends and subset the data:
```{r}
wd <- strptime(date, "%Y-%m-%d")
weekend <- as.factor(weekdays(wd) == "Saturday" | weekdays(wd) == "Sunday")
steps1 <- subset(stepsi, weekend==FALSE)
steps2 <- subset(stepsi, weekend==TRUE)
```
Then we convert the within-day time-stamp to decimal and subset that:
```{r}
inter <- floor(interval/100) + (interval %% 100)/60
int1 <- subset(inter, weekend==FALSE)
int2 <- subset(inter, weekend==TRUE)
```

Then we plot the two series, aggregating over the time-stamp:
```{r}
par(mfrow=c(2,1))

plot(aggregate(steps1 ~ int1, FUN=sum), type="l")
plot(aggregate(steps2 ~ int2, FUN=sum), type="l")
```

As is evident, weekdays are busier, with an earlier start, a distinct
peak at breakfast/commute time, and a generally higher pattern
throughout the rest of the day. 
